---
title: "Forecast using a Neural Network model"
author: "Lorena Romeo"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

# Purpose:
This is the bonus part, in which I will forecast the power consumption using a 
Neural Network model, taking into account also the Temperature regressor.

The first steps are the same of the main project presentation.

## Load necessary libraries:

```{r, warning = FALSE, message = FALSE}
library(readxl)
library(openxlsx)
library(lubridate)
library(dplyr)
library(ggplot2)
library(forecast)
library(zoo)
library(writexl)
```


## Load file

```{r}
#load xlsx file
data_row = read.xlsx("C:/Users/Utilisateur/OneDrive/Desktop/2023-11-Elec-train.xlsx")
#plot(data_row)
head(data_row) #the first record on timestamp column is not in the good format, I need to convert it in the good one 
#str(data_row)
```

## Work on date conversion

```{r}
convert_timestamp = function(ts) {
  if (grepl("^[0-9]+\\.[0-9]+$", ts)) {
    # Handle Excel serial date format
    excel_date = as.numeric(ts)
    origin = as.Date("1899-12-30")
    date = origin + excel_date
    return(as.POSIXct(date))
  } else {
    # Handle standard datetime format
    return(parse_date_time(ts, orders = "m/d/y H:M", tz = "UTC"))
  }
}

# Apply conversion function to Timestamp column
data_row$Timestamp = sapply(data_row$Timestamp, convert_timestamp)

#the timestamp is numeric now, I need to convert to datetime
# Convert numeric Unix epoch timestamps to POSIXct
data_row$Timestamp = as.POSIXct(data_row$Timestamp, origin = "1970-01-01", tz = "UTC")

#head(data_row)
```


## Rename the variables for easiness

```{r}
#now that the dataset is clean I will rename the variables in a easier way
# Rename columns to valid names
elec = data_row %>%
  rename(
    Power_kW = `Power.(kW)`,
    Temp_C = `Temp.(C°)` # Use the exact name here
  )

#verify results
# str(elec)
# head(elec)
```


```{r}
elec <- elec %>%
  filter(!is.na(Power_kW))

#interpolation
elec <- elec %>%
  mutate(
    Power_kW = na.approx(ifelse(Power_kW == 0, NA, Power_kW))
  )
```



## Create train and test dataset

```{r}
# Set up the start and end dates fro train and test dataset
train_start = as.POSIXct("2010-01-01 01:15:00", tz = "UTC")
train_end = as.POSIXct("2010-02-18 23:45:00", tz = "UTC")
test_start = as.POSIXct("2010-02-19 00:00:00",  tz = "UTC")
test_end = as.POSIXct("2010-02-20 23:45:00",  tz = "UTC")

# Create training and test datasets
set.seed(123)
train_data = elec %>% filter(Timestamp >= train_start & Timestamp <= train_end)
test_data = elec %>% filter(Timestamp >= test_start & Timestamp <= test_end)


# str(train_data)
# str(test_data)

#check the boundaries of the datasets
# min(train_data$Timestamp)
# max(train_data$Timestamp)  
# min(test_data$Timestamp) 
# max(test_data$Timestamp)
```

## Create Time Series objects

```{r}
# Create time series objects (without temperature)
# Create time series objects
frequency = 96  # 96 intervals per day (15-minute intervals)
power_train = ts(train_data$Power_kW, start = c(1, 6), end = c(49, 96), frequency = frequency)
#I do not put the end in the test part because the boundary is already set in the code above
power_test = ts(test_data$Power_kW, start = c(50, 1),  frequency = frequency)

# #verify results
# str(power_train)
# str(power_test)
```



# Create a first Neural Network Model using NNAR

This is a first attempt on creating a Neural Network model without fine tuning.

```{r}
#create neural model
fit_nnar = nnetar(power_train, xreg = train_data$Temp_C)

#check predictions using test data
prev_nnar = forecast(fit_nnar, xreg = test_data$Temp_C, h = length(power_test))
```


## Check quality of the model

```{r}
# # Mean Absolute Error
# mae = mean(abs(prev_nnar$mean - power_test))
# print(mae)
# 
# # Mean Squared Error
# mse = mean((prev_nnar$mean - power_test)^2)
# print(mse)

# Root Mean Squared Error
nnr1_rmse = sqrt(mean((prev_nnar$mean - power_test)^2))
print(nnr1_rmse)

# # Mean Absolute Percentage Error
# mape = mean(abs((prev_nnar$mean - power_test) / power_test)) * 100
# print(mape)
# 
# # Symmetric Mean Absolute Percentage Error
# smape = mean(2 * abs(prev_nnar$mean - power_test) / (abs(prev_nnar$mean) + abs(power_test))) * 100
# print(smape)

```
From the RMSE I can suppose that with a bit of fine tune the model can have a better fit.


# Retrieve the parameters of the auto generated model to understand how to fine tune it

I extract the parameters of the auto generated nnar and then decide what to change
```{r}
# View specific model components
p_value = fit_nnar$p  # Number of non-seasonal lags
P_value = fit_nnar$P  # Number of seasonal lags
size_value = fit_nnar$size  # Number of neurons in the hidden layer
lambda_value = fit_nnar$lambda  # Box-Cox transformation parameter (if used)
repeats_value = fit_nnar$repeats  # Number of networks averaged


# Print these values
cat("Number of lags (p):", p_value, "\n")
cat("Number of seasonal lags (P):", P_value, "\n")
cat("Number of neurons (size):", size_value, "\n")
cat("Box-Cox transformation (lambda):", lambda_value, "\n")
cat("Number of repeats:", repeats_value, "\n")
```

#Fit manually a new model and forecast

I can now try to fine tune the NN model.

```{r}
#1st attempt
fit2_nnar = nnetar(
  power_train,
  p = 15,
  P = 1,
  size = 15,
  repeats = 10,
  lambda = "auto",
  xreg = train_data$Temp_C
)
```


# Make forecast based on test set and check quality of the model.

```{r}
# Make forecasts
prev_nnar2 = forecast(fit2_nnar, xreg = test_data$Temp_C, h = length(power_test))


# Evaluate model performance on adjusted test data
mse = mean((prev_nnar2$mean - power_test)^2)
print(mse)

# Root Mean Squared Error
nnr2_rmse = sqrt(mean((prev_nnar2$mean - power_test)^2))
print(nnr2_rmse)
```


# Fit manually another model and forecast
```{r}
#2nd attempt
fit3_nnar = nnetar(
  power_train,
  p = 15,
  P = 3, 
  size = 15, 
  repeats = 10,
  lambda = "auto",
  xreg = train_data$Temp_C
)
```


# Make forecast based on test set and check quality of the model.
```{r}
# Make forecasts
prev_nnar3 = forecast(fit3_nnar, xreg = test_data$Temp_C, h = length(power_test))

# Evaluate model performance on adjusted test data
mse = mean((prev_nnar3$mean - power_test)^2)
print(mse)

# Root Mean Squared Error
nnr3_rmse = sqrt(mean((prev_nnar3$mean - power_test)^2))
print(nnr3_rmse)
```

From the RMSE of the three NN models I can see that the 2nd attempt (nnr2) is the
best one. 
I will use it to forecast the power consumption on the 21st February.


# Plot the best NN model with the test data

```{r}
#to see in details, plot only forecast
plot(prev_nnar2$mean, type = "l", col = "blue", lwd = 2, 
     xlab= "Time", ylab = "Power (kW)", main = "Forecats Vs Actuals")
#and add the actual (test values)
lines(power_test, col = "red", lwd = 1, lty = 2)
#add the legend
legend("topleft", legend = c("Forecast","Actual"),
       col = c("blue", "red"), lty = c(1,2), lwd = c(1,2))
```


# Prepare to Forecast the Power kw of the 21st February

```{r}
# I need to create a new ts object with the complete dataset 

#set up the start and end dates for the complete dataset until 20th feb
start = as.POSIXct("2010-01-01 01:15:00", tz = "UTC")
end = as.POSIXct("2010-02-20 23:45:00",  tz = "UTC")

#create complete datasets until 20 feb 2010
complete_data = elec %>% filter(Timestamp >= start & Timestamp <= end)

#create time series objects (without temperature)
frequency = 96  # 96 intervals per day (15-minute intervals)
power_complete = ts(complete_data$Power_kW, start = c(1, 6), frequency = frequency)

#then I create a ts for the complete temperature 
temp_next = ts(complete_data$Temp_C, start = c(1, 6), frequency = frequency)

#extract the relevant temperature series for the next 96 time points
temp_forecast = tail(complete_data$Temp_C, 96)  # Extract last 96 temperature observations

#I create a ts object for the temperature of the 21 Feb
temp_forecast_ts = ts(temp_forecast, frequency = 96)
```


# Train the model using the complete dataset

```{r}
#I train the model using complete power and temperature data (until 20 Feb 2010)
naar_complete = nnetar(
  power_complete,
  p = 15,
  P = 1, 
  size = 15, 
  repeats = 10,
  lambda = "auto",
  xreg = complete_data$Temp_C
)

```


# Forecast the next day: 21st February

```{r}
#forecast the next 96 time points
nnar_next = forecast(naar_complete, h = 96, xreg = temp_forecast_ts)

```

# Plot the Forecast together wtih the complete dataset

```{r}
# Plot the complete data and forecast 
autoplot(power_complete) +
   autolayer(nnar_next$mean, series = "Forecast of 21st Feb with Neural Network", color = "red") +
   ggtitle("Forecats in red the Power_kw demanded on 21st February 2010") +
   ylab("Power (kW)") +
   xlab("Time") 
```

# Eventually print the Forecast results

```{r}
print(nnar_next)
```

# Export the results in excel
```{r}
#correctly convert forecast power to a data frame
forecast_nnr = as.data.frame(nnar_next)
#extracting just the forecast points
point_forecast_nnr = forecast_nnr[["Point Forecast"]]
#create columns that I will use as 1st column in excel file
nnr_forecast_withtemp = as.data.frame(point_forecast_nnr)


#rename the columns to more descriptive names
colnames(nnr_forecast_withtemp) = c("NNR_With_Temperature")
#save the data frames to an Excel file
write_xlsx(nnr_forecast_withtemp, path = "C:/Users/Utilisateur/OneDrive/Desktop/Romeo_Lorena_NNR_secondsub.xlsx")
```

